Model Selection to Find a Sufficient Model Fitting to Thermal Performance of Metabolic Traits\footnotemark[1]

### 2021-01-19 MiniProject/Code/Report.tex
CMEECourseWork

\title{Model Selection to Find a Most Sufficient Model Fitting to Thermal Performance of Metabolic Traits\footnotemark[1]}

@@

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[1]{Word Count: 2611}
\footnotetext[2]{Department of Life Sciences (Silwood Park), Imperial College of Science, Technology and Medicine}

@@

  \linenumbers
  \begin{abstract}
    Determining thermal dependence of biological traits provides better understanding in distribution patterns of species 
    and is helpful in assessing the impact of climate change on biological systems. Here, I fitted quadratic polynomial 
    model, cubic polynomial model, Briere model and Ratkowsky model to observed data of metabolic traits under different 
    temperatures to calculate thermal performance curves. Model averaging and selection by either AIC or BIC were 
    conducted to find one most sufficient model fitting to empirical data of each curve. 
    In most curves, AIC and BIC give support to 
    the same model and Ratkowsky model has the highest frequency of being the most suitable model 
    under two selection criteria. However, for some thermal performance curves in the dataset, even the best model 
    selected here fail in describing thermal dependence of metabolic traits, indicating that the four plausible models 
    are not suitable and more models need to be involved. Overall, the results show that Ratkowsky model is sufficient 
    for most thermal performance curves involved in this study.

@@

  Temperature is an important abiotic factor that influences the functioning of organisms, especially 
  for ectotherms whose body temperatures reflect that of their environments to different extent. It affects 
  the rates of biochemical reactions and consequently has profound influence on different biological 
  activities like metabolism, growth, development and reproduction \cite{sinclair2016can}. And this can 
  lead to further consequences in abundance and distribution of organisms \cite{schulte2011thermal}. 
  Besides, interactions among species are also affected by thermal environment \cite{grigaltchik2012thermal}.  
  \newline
  Thermal performance curve describes the dependence of performance (usually, the rate of a physiological 
  activities or fitness) on temperature quantitatively \cite{sinclair2016can}. Generally, organisms response 
  to temperature fluctuation in a similar manner. The value of performance peaks at some temperature point, 
  and decreases when temperature deviates from the maximum point, while extremely high or low temperatures are lethal 
  \cite{sinclair2016can,krenek2011thermal}. 
  The mathematical modelling of thermal performance curve provides a framework for the estimation of 
  physiological characters like the optimum temperature at which performance peaks, upper and lower 
  thermal limit thermal performance, and thermal tolerance range \cite{krenek2011thermal}. 
  Besides, it is also helpful in predicting 
  performance of organisms when temperature fluctuates. Thus, thermal performance curve is helpful in better 
  understanding distribution patterns of species. There is evidence that 
  the thermal tolerance range of a species is often corresponded 
  to the temperature fluctuation of its habitat, indicating the 
  importance of temperature as a selective force in shaping distribution pattern \cite{sunday2011global, clusella2011climatic}. 
  There is also evidence showing 
  that thermal sensitivity plays a role in the dynamics of biodiversity \cite{allen2002global}. Furthermore, 
  investigations in thermal performance curve help better predict the impact of current climate change on 
  biological process \cite{krenek2011thermal, sears2011introduction}.  
  \newline
  Currently, there are numerous mathematical models with different properties  
  that are used to illustrate thermal performance curves of different clades 
  \cite{krenek2011thermal,angilletta2006estimating}. Which model, or models are sufficient in describing thermal dependence 
  of performance is the main concern of this study. 
  Using a dataset of observed data from 841 thermal performance curves, 
  I calculated these curves by fitting 
  four plausible models to them. 
  Then I conducted model averaging and selection to find one model that is most sufficient to describe the observed data.
   Akaike information criterion (AIC) or 
  Bayesian information criterion (BIC) was used as model selection criterion \cite{johnson2004model}. 
  Besides, whether metabolic trait or habitat type is corresponded to models that get most support from 
  empirical data was assessed.

  \section{Materials and Methods}
    \subsection{Original Data}
    The original dataset used here is named as ThermRespData.csv and its field names are defined in another 
    file called BiotraitsTemplateDescription.pdf. Both files are accessible in 
    \newline 
    \href{https://github.com/mhasoba/TheMulQuaBio/tree/master/content/data}{https://github.com/mhasoba/TheMulQuaBio/tree/master/content/data}. 
    \newline 
    The main fields of interest are OriginalTraitValue (trait values) and ConTemp (the temperatures 
    in Celsius scale). Each thermal performance curve is corresponded to a unique ID from 1 to 903. 
    Among them 62 curves contain negative trait values and were removed. 
    So a total of 841 thermal performance curves was used in following analysis.

  \subsection{Plausible Models} 
    Denote by \textit{B} the trait value and \textit{T} the temperature. The function that describes 
    the relationship between temperature and trait value is  
    \begin{equation} 
      B = B(T) 
    \end{equation} 
    Plausible models are: (1) quadratic polynomial model with constants  
    $B_0$, $B_1$ and $B_2$: 
    \begin{equation} 
    B(T) = B_0 + B_1 T + B_2 T^2 
    \end{equation} 
    (2) cubic polynomial model with constants  
    $B_0$, $B_1$, $B_2$ and $B_3$: 
    \begin{equation} 
    B(T) = B_0 + B_1 T + B_2 T^2 + B_3 T^3 
    \end{equation} 
    (3) Briere model with constants  
    $B_0$, $T_0$ and $T_m$ 
    \cite{briere1999novel}: 
    \begin{equation}\label{equ:Briere} 
    B(T) = B_0 T (T - T_0) \sqrt{T_m - T} 
    \end{equation} 
    $T_0$ and $T_m$ are lower and upper thermal limit thermal performance. 
    \newline 
    (4) Ratkowsky model with constants  
    $T_0$, $T_m$, $a$ and $b$  
    \cite{ratkowsky1983model}: 
    \begin{equation} 
    B(T) = \{a (T-T_0) (1-e^{b(T-T_m)})\}^2 
    \end{equation} 
    $T_0$ and $T_m$ are lower and higher thermal limit thermal performance. 
  \subsection{Model Fitting} 
    To conduct model fitting, least square method was used, and non-linear models were fitted using Levenberg-Marquardt  
    algorithm \cite{ranganathan2004levenberg, lourakis2005brief}. Maximum iteration number was 200.  
    \newline 
    For fitting of Briere model, start values of $T_0$ and $T_m$ were the lowest and highest sampled temperatures 
    of each curve and they are denoted by $T_{0st}$ and $T_{mst}$ respectively. From Eq. (\ref{equ:Briere}), 
    \begin{equation}\label{lalala} 
    B_0 = \frac{B(T)}{T (T - T_0) \sqrt{T_m - T}} 
    \end{equation} 
    For each curve, remove samples with lowest and highest temperatures and denote the rest as  
    $\{(T_i, B_i), i = 1,2,\cdots,n-2\}$, where $n$ is sample size.  
    So from Eq. (\ref{lalala}), a rough estimation of $B_0$, denoted by $B_{0est}$ can be given by  
    \begin{equation}\label{esti} 
    B_{0est} = \frac{\sum_{i=1}^{n-2}\frac{B_i}{T_i (T_i - T_{0st}) \sqrt{T_{mst} - T_i}}}{n-2} 
    \end{equation} 
    The start values of $B_0$ was taken from a uniform distribution between $0$ and $2B_{0est}$. 
    Each curve was fitted 100 times with different start values of $B_0$. 
    \newline 
    As for fitting of Ratkowsky model, start values of $T_0$ and $T_m$ were the lowest and highest sampled temperatures 
    of each curve. Start values of $a$ and $b$ were taken from set  
    $\{10^i: i = -5,-4,-3,-2,-1,0,1,2,3,4\}$. So for each curve, there are ten start values of $a$ and $b$,  
    and 100 combinations of them. Hence, each curve was fitted 100 times with different start values.

@@

        \subsection{Assess Best Models for Curves of Different Metabolic or Habitat Types}
        To assess preferred models for thermal performance curves of differnet metabolic traits or species living in different 
        habitat types, 18 curves were removed because AIC and BIC supported different models. So a total of 823 
        curves were involved.
        \newline  
        In the original dataset, the metabolic trait of each curve is defined in the field of StandardisedTraitName 
        (trait name for comparison), 
        which contains three categories: gross photosynthesis rate, net photosynthesis rate and respiration rate. Another field 
        relevant to 
        trait name is OriginalTraitName (trait name as it found in source). It should be noted there are four curves 
        have no StandardisedTraitName, while their OriginalTraitName are oxygen evolution 
        rate in thylakoid membranes (IDs: 698, 699) and cell specific photosynthesis rate (IDs: 441, 442). They were not 
        involved in assessing which model is suitable for different metabolic traits.
        As for habit type of each curve, it is defined in another field (Habitat), which is composed of four types: aquatic, 
        freshwater, freshwater/terrestrial, marine and terrestrial. Here I assessed which model is suitable for determining 
        thermal performance of species living in terrestrial or non-terrestrial (aquatic, 
        freshwater, freshwater/terrestrial and marine) habitats.
        \newline  
        Table 2 illustrates the sufficient models for thermal performance curves of different metabolic traits. 
        A total of 819 thermal performance curves of three metabolic traits are involved. 
        For models of each metabolic trait, non-linear  
        model occupies a larger proportion 
        (75.0\% of gross photosynthesis rate, 78.8\% of net photosynthesis rate, and 66.2\% of respiration rate) than linear 
        models. Besides, Ratkowsky model takes a bigger amount than Briere model in all metabolic traits except gross 
        photosynthesis rate.
        \newline
        Then I classified thermal performance curves according to whether the species studied is specific in terrestrial 
        habitats (Table 3). Non-linear model 
        takes the biggest amount of sufficient models for both two kinds of habitat type 
        (82.1\% for terrestrial habitat and 55.2\% of non-terrestrial habitat), and Ratkowsky model occupies biggest proportion 
        among four models. 
        \begin{table}[H]
          \centering
          \caption{Selected Models for Thermal Performance Curves of Different Metabolic Traits}
          \begin{tabular}{@{}ccccc@{}}
          \toprule
          Model     & Gross photosynthesis rate & Net photosynthesis rate & Respiration rate  & Total \\ \midrule
          Quadratic & 0   & 19  & 36  & 55    \\
          Cubic     & 9   & 80  & 71  & 160   \\
          Briere    & 16  & 73  & 94  & 183   \\
          Ratkowsky & 11  & 294 & 116 & 421   \\
          Total     & 36  & 466 & 317 & 819   \\ \bottomrule
          \end{tabular}
          \end{table}

@@

    \subsection{Preferred Models May not Describe the Observed Data Well}
    Through model selection, I selected one model for each curve and it is the most sufficient one in terms of describing 
    the thermal performance curve among four plausible models. However, the preferred models of some curves still 
    fail in describing thermal performance throughout the temperature 
    range between lower and upper thermal limits of performance. As the growth of temperature in this range, 
    performance is supposed to increases, peaks at some point, and then drops \cite{sinclair2016can,krenek2011thermal}. 
    The preferred models of some curves are 
    not bell-shaped (Figure. 3). Furthermore, ecologically unrealistic estimations 
    of thermal limits of performance are also given by some selected models (Figure 4).

@@

  is best supported by empirical data and is able to provide relatively precise estimations of controlled variable, 

@@

  approximation reflecting the probability that each plausible model is the true model that generated the empirical 

@@

  the existence of "true model" is problematic \cite{box1979all}, and the theoretical basis of BIC is 
  controversial \cite{kuha2004aic}. As for AIC, it measures the predictive performance of 

@@

  be realistic, as exemplified in Figure 4. Such models may fit to the observed data well (Figure 3d and Figure 4a), but 

>>>>

\title{Model Selection to Find a Sufficient Model Fitting to Thermal Performance of Metabolic Traits\footnotemark[1]}

@@

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[1]{Word Count: 2597}
\footnotetext[2]{Department of Life Sciences (Silwood Park), Imperial College of Science, Technology and Medicine}

@@

  \linenumbers
  \begin{abstract}
    Determining thermal dependence of biological traits provides understanding in distribution patterns of species 
    and is helpful in assessing the impact of climate change on biological systems. Here, I fitted 
    four plausibe models to observed data of metabolic traits under different 
    temperatures to calculate thermal performance curves. Model selection by either AIC or BIC were 
    conducted to find one most sufficient model fitting to empirical data of each curve. 
    In most curves, AIC and BIC give support to 
    the same model and Ratkowsky model has the highest frequency of being selected. 
    However, for some thermal performance curves in the dataset, the selected model 
    fails in describing thermal dependence of metabolic traits, indicating that the four plausible models 
    are not suitable and more models need to be involved. Overall, the results show that Ratkowsky model is sufficient 
    for most thermal performance curves involved in this study.

@@

  Temperature is an important abiotic factor that influences the functioning of organisms, especially 
  for ectotherms whose body temperatures reflect that of their environments to different extent. It affects 
  the rates of biochemical reactions and consequently has profound influence on different biological 
  activities like metabolism, growth, development and reproduction \cite{sinclair2016can}. And this can 
  lead to further consequences in abundance and distribution of organisms \cite{schulte2011thermal}. 
  Besides, interactions among species are also affected by thermal environment \cite{grigaltchik2012thermal}.  
  \newline
  Thermal performance curve describes the dependence of performance (usually, the rate of a physiological 
  activities or fitness) on temperature quantitatively \cite{sinclair2016can}. Generally, organisms response 
  to temperature fluctuation in a similar manner. The value of performance peaks at some temperature point 
  and decreases when temperature deviates from the maximum point, while extremely high or low temperatures are lethal 
  \cite{sinclair2016can,krenek2011thermal}. 
  The mathematical modelling of thermal performance curve provides a framework for the estimation of 
  physiological characters like the optimum temperature at which performance peaks, upper and lower 
  thermal limits of performance, and thermal tolerance range \cite{krenek2011thermal}. 
  Besides, it is also helpful in predicting 
  performance of organisms when temperature fluctuates. Thus, thermal performance curve is helpful in better 
  understanding distribution patterns of species. There is evidence that 
  the thermal tolerance range of a species is often corresponded 
  to the temperature fluctuation of its habitat, indicating the 
  importance of temperature as a selective force in shaping distribution pattern \cite{sunday2011global, clusella2011climatic}. 
  There is also evidence showing 
  that thermal sensitivity plays a role in the dynamics of biodiversity \cite{allen2002global}. Furthermore, 
  investigations in thermal performance curve help better predict the impact of current climate change on 
  biological process \cite{krenek2011thermal, sears2011introduction}.  
  \newline
  Currently, there are numerous mathematical models with different properties  
  that are used to illustrate thermal performance curves of different clades 
  \cite{krenek2011thermal,angilletta2006estimating}. Which model, or models are sufficient in describing thermal dependence 
  of performance is the main concern of this study. 
  Using a dataset of observed data from 841 thermal performance curves of different metabolic traits, 
  I calculated these curves by fitting 
  four plausible models to them. 
  Then I conducted model selection to find one model that is most sufficient to describe the observed data. 
  Akaike information criterion (AIC) or 
  Bayesian information criterion (BIC) was used as model selection criterion \cite{johnson2004model}. 
  Besides, whether metabolic trait or habitat type is corresponded to models that get most support from 
  empirical data was assessed.

  \section{Materials and Methods}
    \subsection{Original Data}
    The original dataset used here is named as ThermRespData.csv and its field names are defined in another 
    file called BiotraitsTemplateDescription.pdf. Both files are accessible in 
    \newline 
    \href{https://github.com/mhasoba/TheMulQuaBio/tree/master/content/data}{https://github.com/mhasoba/TheMulQuaBio/tree/master/content/data}. 
    \newline 
    The main fields of interest are OriginalTraitValue (trait values) and ConTemp (temperatures 
    in Celsius scale). Each thermal performance curve is corresponded to a unique ID from 1 to 903. 
    Among them 62 curves contain negative trait values and were removed. 
    So a total of 841 thermal performance curves was used in following analysis.

  \subsection{Plausible Models} 
    Denote by \textit{B} the trait value and \textit{T} the temperature. The function that describes 
    the relationship between temperature and trait value is  
    \begin{equation} 
      B = B(T) 
    \end{equation} 
    Plausible models are: (1) quadratic polynomial model with constants  
    $B_0$, $B_1$ and $B_2$: 
    \begin{equation} 
    B(T) = B_0 + B_1 T + B_2 T^2 
    \end{equation} 
    (2) cubic polynomial model with constants  
    $B_0$, $B_1$, $B_2$ and $B_3$: 
    \begin{equation} 
    B(T) = B_0 + B_1 T + B_2 T^2 + B_3 T^3 
    \end{equation} 
    (3) Briere model with constants  
    $B_0$, $T_0$ and $T_m$ 
    \cite{briere1999novel}: 
    \begin{equation}\label{equ:Briere} 
    B(T) = B_0 T (T - T_0) \sqrt{T_m - T} 
    \end{equation} 
    $T_0$ and $T_m$ are lower and upper thermal limits of performance. 
    \newline 
    (4) Ratkowsky model with constants  
    $T_0$, $T_m$, $a$ and $b$  
    \cite{ratkowsky1983model}: 
    \begin{equation} 
    B(T) = \{a (T-T_0) (1-e^{b(T-T_m)})\}^2 
    \end{equation} 
    $T_0$ and $T_m$ are lower and higher thermal limits of performance. 
  \subsection{Model Fitting} 
    To conduct model fitting, least square method was used, and non-linear models were fitted using Levenberg-Marquardt  
    algorithm \cite{ranganathan2004levenberg, lourakis2005brief}. Maximum iteration number was 200.  
    \newline 
    For fitting of Briere model, start values of $T_0$ and $T_m$ were the lowest and highest sampled temperatures 
    of each curve and they are denoted by $T_{0st}$ and $T_{mst}$ respectively. From Eq. (\ref{equ:Briere}), 
    \begin{equation}\label{lalala} 
    B_0 = \frac{B(T)}{T (T - T_0) \sqrt{T_m - T}} 
    \end{equation} 
    For each curve, remove samples with lowest and highest temperatures and denote the rest as  
    $\{(T_i, B_i), i = 1,2,\cdots,n-2\}$, where $n$ is sample size.  
    So from Eq. (\ref{lalala}), a rough estimation of $B_0$, denoted by $B_{0est}$ can be given by  
    \begin{equation}\label{esti} 
    B_{0est} = \frac{\sum_{i=1}^{n-2}\frac{B_i}{T_i (T_i - T_{0st}) \sqrt{T_{mst} - T_i}}}{n-2} 
    \end{equation} 
    The start values of $B_0$ was taken from a uniform distribution between $0$ and $2B_{0est}$. 
    Each curve was fitted 100 times with different start values of $B_0$. 
    \newline 
    As for fitting of Ratkowsky model, start values of $T_0$ and $T_m$ were the lowest and highest sampled temperatures 
    of each curve. Start values of $a$ and $b$ were taken from set  
    $\{10^i: i = -5,-4,-3,-2,-1,0,1,2,3,4\}$ iteratively. So for each curve, there are ten start values of $a$ and $b$,  
    and 100 combinations of them. Hence, each curve was fitted 100 times with different start values.

@@

        \subsection{Selected Models for Curves of Different Metabolic or Habitat Types}
        To assess preferred models for thermal performance curves of differnet metabolic traits or species living in different 
        habitat types, 18 curves were removed because AIC and BIC supported different models. So a total of 823 
        curves were involved.
        \newline  
        In the original dataset, the metabolic trait of each curve is defined in the field of StandardisedTraitName 
        (trait name for comparison), 
        which contains three categories: gross photosynthesis rate, net photosynthesis rate and respiration rate. Another field 
        relevant to 
        trait name is OriginalTraitName (trait name as it found in source). It should be noted there are four curves 
        have no StandardisedTraitName, while their OriginalTraitName are oxygen evolution 
        rate in thylakoid membranes (IDs: 698, 699) and cell specific photosynthesis rate (IDs: 441, 442). They were not 
        involved in assessing which model is suitable for different metabolic traits.
        As for habit type of each curve, it is defined in another field (Habitat), which is composed of four types: aquatic, 
        freshwater, freshwater/terrestrial, marine and terrestrial. Here I assessed which model is suitable for determining 
        thermal performance of species living in terrestrial or non-terrestrial (aquatic, 
        freshwater, freshwater/terrestrial and marine) habitats.
        \newline  
        Table 2 illustrates the sufficient models for thermal performance curves of different metabolic traits. 
        A total of 819 thermal performance curves of three metabolic traits are involved. 
        For models of each metabolic trait, non-linear  
        model occupies a larger proportion 
        (75.0\% in gross photosynthesis rate, 78.8\% in net photosynthesis rate, and 66.2\% in respiration rate) than linear 
        models. Besides, Ratkowsky model takes a bigger amount than Briere model in all metabolic traits except gross 
        photosynthesis rate.
        \newline
        Then I classified thermal performance curves according to whether the species studied is specific in terrestrial 
        habitats (Table 3). Non-linear model 
        takes the biggest amount of sufficient models for both two kinds of habitat type 
        (82.1\% for terrestrial habitat and 55.2\% of non-terrestrial habitat), and Ratkowsky model occupies biggest proportion 
        among four plausible models. 
        \begin{table}[H]
          \centering
          \caption{Selected Models for Thermal Performance Curves of Different Metabolic Traits}
          \begin{tabular}{@{}ccccc@{}}
          \toprule
          Model     & Gross photosynthesis rate & Net photosynthesis rate & Respiration rate  & Total \\ \midrule
          Quadratic & 0   & 19  & 36  & 55    \\
          Cubic     & 9   & 80  & 71  & 160   \\
          Briere    & 16  & 73  & 94  & 183   \\
          Ratkowsky & 11  & 294 & 116 & 421   \\
          Total     & 36  & 466 & 317 & 819   \\ \bottomrule
          \end{tabular}
          \end{table}

@@

    \subsection{Preferred Models May not Describe the Observed Data Sufficiently}
    Through model selection, I selected one model for each curve and it is the most sufficient one in terms of describing 
    the thermal performance curve among four plausible models. However, the preferred models of some curves still 
    fail in describing thermal performance throughout the temperature 
    range between lower and upper thermal limits of performance. As the growth of temperature in this range, 
    performance is supposed to increases, peaks at some point, and then drops \cite{sinclair2016can,krenek2011thermal}. 
    The preferred models of some curves are 
    not bell-shaped (Figure. 3). Furthermore, ecologically unrealistic estimations 
    of thermal limits of performance are also given by some selected models (Figure 4).

@@

  is best supported by empirical data and is able to provide relatively precise estimations of dependent variable using 

@@

  approximation reflecting the probability that the plausible model is the true model that generated the empirical 

@@

  whether there is such a "true model" is controversial \cite{box1979all}, and the theoretical basis of BIC is 
  disputed \cite{kuha2004aic}. As for AIC, it measures the predictive performance of 

@@

  be realistic, as exemplified in Figure 4. Such models may fit to the observed data well (\textit{e.g.} Figure 3d and Figure 4a), but 
